A linear inside-outside algorithm for correcting sequencing errors in structured RNAs

### 2012-10-07 Recomb/methods_RECOMB.tex
New title + removed todo

\subsubsection{Energy contribution}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is the tabulated free-energy of a stacking pair in the Turner model if available, $0$ if $ab=\varnothing$ (no base-pair to stack onto), and using default value $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\text{GU},\text{UG},\text{CG},\text{GC}, \text{AU or UA}\}$). This latter parameter allows to choose whether to completely forbid a sequence inducing invalid base pairs ($\beta = \infty$), or to simply penalize it.
The imprecision introduced by this simplification of the Turner model remains reasonable, since the targeted secondary structure is fixed 
(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsection{isostericity contribution}
The concept of isostericity captures the geometric superimposability of two base-pairs, using individual scores given by Stombaugh~\emph{et al}~\cite{Stombaugh2009} based on a probabilistic model. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gapless RNA alignment $\Omega$,  the contribution of the isostericity to the pseudo-energy is given by
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence in comparison with the reference alignment.
The $\ISO$ function uses the \todo[fancyline, inline]{\relsize{-1}Is this true? If so, we might want to try taking the most likely edge/orientation combination for each couple instead.}{Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, with $0$ being assigned to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the alignment sequences.

@@

Let $s$ an RNA 
sequence and $m\geq 0$. $S$ is considered as  one derivation of the \todo[fancyline]{A bit tough: SCFG's were never mentioned before. Maybe introduce this with a sentence relating problem to parsing}{SCFG} generating  
all  secondary structures of length  $|s|$. We are interested in  
the probability of a given position being a specific nucleotide,
over all sequences having at most $m$ mutations from $s$, under the SCFG derivation $S$ 
(formally 
$\mathbb{P}(s_i = x\mid s,\Omega, S,m)$). 
We define as a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990}, allowing us to obtain the
 desired probability,  the two  functions
$\Z{i,j}{m}{a,b}$ and $\Y{i,j}{m}{a,b}$. The former presented in 
Equations~\eqref{eq:Z_in} and~\eqref{eq:Z_rec}, our 
version of the \emph{inside}, computes the partition function between $i$ and $j$ included, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$) and containing 
$m$ mutations. The latter in Equations~\eqref{eq:Y_in} and~\eqref{eq:Y_rec},
 computes the \emph{outside}, in particular,  
the partition function considering only between $[0,i]\cup[j,n-1]$
knowing  that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$) and containing 
 $m$ mutations outside.

@@

We are now left to compute the probability that a \todo[fancyline]{Maybe worth doing it for base-pairs? maybe not\ldots}{given position is a given nucleotide}.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of allowed mutations. 
\begin{align*}
	\mathbb{P}(s_i = x\mid s,\Omega, S,M) &:= \frac{\mathcal{B}(i,x,s,\Omega,S,M)}{\sum_{m=0}^{M}\Z{0,n-1}{m}{X,X}}\\ 
\mathcal{B}(i,x,s,\Omega,S,M)&=
 \left\{
	\begin{array}{ll}
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		&\text{If }S_i = -1\\
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
			\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\Z{i+1,k-1}{m'}{x,b}
		&\text{If }S_i=k>i\\
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-E^{\Omega, \beta}_{(k,i),\varnothing\to bx}}{RT}}
			\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\Z{k+1,i-1}{m'}{b,x}
		&\text{If }S_i=k<i
	\end{array}\right.
\end{align*}

In every case, the denominator is the sum of the partitions function of exactly $m$ mutations, \todo[fancyline,inline]{\relsize{-1}Seems weird to do this for a given upper bound $M$ (as opposed to exact $m$), and it makes the equations a bit more complicated. Anyway, there is roughly $3n$ times more sequences with $m+1$ mutations than with $m$ muts, so I expect the $m+1$ guys are so dominating in the final probabilities that it may not be worth the trouble of combining them with the \emph{little guys} ($\#muts < m$)}{for $m$ smaller or equal to our target $M$}. The numerators are divided in the following three cases.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, we are concerned by the weights
over all sequences which have at position $i$ nucleotide $x$, which is exactly the sum
of the values of $\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}$, for all $m$ between $0$ and $M$.
\item[$S_i=k>i$:] Since we need to respect the derivation of the secondary structure $S$, if 
position $i$ is paired, we must consider the two partition functions. The \emph{outside} of the 
base pair, and the \emph{inside}, for all possible values for the nucleotide at position $k$, and
all possible distribution of the mutant positions between the inside and outside of the base pair. We also add the term of isostericity for this specific base pair.
\item[$S_i=k<i$:] Same as above, but with position $i$ pairing with a lower position.
\end{description}

>>>>

\subsubsection{Energy contribution}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is the tabulated free-energy of a stacking pair in the Turner model if available, $0$ if $ab=\varnothing$ (no base-pair to stack onto), and using default value $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\text{GU},\text{UG},\text{CG},\text{GC}, \text{AU or UA}\}$). This latter parameter allows to choose whether to completely forbid a sequence inducing invalid base pairs ($\beta = \infty$), or to simply penalize it.
The imprecision introduced by this simplification of the Turner model remains reasonable, since the targeted secondary structure is fixed 
(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsection{isostericity contribution}
The concept of isostericity captures the geometric superimposability of two base-pairs, using individual scores given by Stombaugh~\emph{et al}~\cite{Stombaugh2009} based on a probabilistic model. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gapless RNA alignment $\Omega$,  the contribution of the isostericity to the pseudo-energy is given by
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence in comparison with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, with $0$ being assigned to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the alignment sequences.

@@

Let $s$ an RNA 
sequence and $m\geq 0$. $S$ is considered as  one derivation of the 
{SCFG} generating  
all  secondary structures of length  $|s|$. We are interested in  
the probability of a given position being a specific nucleotide,
over all sequences having at most $m$ mutations from $s$, under the SCFG derivation $S$ 
(formally 
$\mathbb{P}(s_i = x\mid s,\Omega, S,m)$). 
We define as a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990}, allowing us to obtain the
 desired probability,  the two  functions
$\Z{i,j}{m}{a,b}$ and $\Y{i,j}{m}{a,b}$. The former presented in 
Equations~\eqref{eq:Z_in} and~\eqref{eq:Z_rec}, our 
version of the \emph{inside}, computes the partition function between $i$ and $j$ included, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$) and containing 
$m$ mutations. The latter in Equations~\eqref{eq:Y_in} and~\eqref{eq:Y_rec},
 computes the \emph{outside}, in particular,  
the partition function considering only between $[0,i]\cup[j,n-1]$
knowing  that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$) and containing 
 $m$ mutations outside.

@@

We are now left to compute the probability that a given position is a given nucleotide.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of allowed mutations. 
\begin{align*}
	\mathbb{P}(s_i = x\mid s,\Omega, S,M) &:= \frac{\mathcal{B}(i,x,s,\Omega,S,M)}{\sum_{m=0}^{M}\Z{0,n-1}{m}{X,X}}\\ 
\mathcal{B}(i,x,s,\Omega,S,M)&=
 \left\{
	\begin{array}{ll}
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		&\text{If }S_i = -1\\
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
			\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\Z{i+1,k-1}{m'}{x,b}
		&\text{If }S_i=k>i\\
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-E^{\Omega, \beta}_{(k,i),\varnothing\to bx}}{RT}}
			\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\Z{k+1,i-1}{m'}{b,x}
		&\text{If }S_i=k<i
	\end{array}\right.
\end{align*}

In every case, the denominator is the sum of the partitions function of exactly $m$ mutations, 
for $m$ smaller or equal to our target $M$. The numerators are divided in the following three cases.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, we are concerned by the weights
over all sequences which have at position $i$ nucleotide $x$, which is exactly the sum
of the values of $\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}$, for all $m$ between $0$ and $M$.
\item[$S_i=k>i$:] Since we need to respect the derivation of the secondary structure $S$, if 
position $i$ is paired, we must consider the two partition functions. The \emph{outside} of the 
base pair, and the \emph{inside}, for all possible values for the nucleotide at position $k$, and
all possible distribution of the mutant positions between the inside and outside of the base pair. We also add the term of isostericity for this specific base pair.
\item[$S_i=k<i$:] Same as above, but with position $i$ pairing with a lower position.
\end{description}
